An evaluation framework for trajectory inference methods

### 2017-11-21 analysis/paper/paper.Rmd
updated introduction

---
title: "An evaluation framework for trajectory inference methods"
author:
  - Robrecht Cannoodt*
  - Wouter Saelens*
  - Helena Todorov
  - Yvan Saeys
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
    fig_caption: yes
    reference_docx: templates/template.docx
bibliography: references.bib
csl: nature-biotechnology.csl
abstract: |
  Using single-cell transcriptomics data, it is now possible to computationally order cells along trajectories, allowing the unbiased transcriptome-wide study of cellular dynamic processes. Dozens of trajectory inference methods have been developed, early methods mostly focussing on linear trajectories, while recently methods also model more complex branching trajectories. However, the performance of these methods have not yet been comprehensively evaluated. Here, we compare a total of 22 trajectory inference methods, on both realistic synthetic datasets and a large set is real datasets. We employ a novel parameter optimisation procedure, and compare methods using several metrics, including accuracy of the inferred ordering, correctness of the network wiring, code quality and user friendliness. We conclude with practical guidelines for method users, and also believe our evaluation framework can be used to spearhead the development for new methods. Our whole evaluation pipeline is fully reproducible, can be easily extended and is available at www.github.com/dynverse.
keywords: |
  trajectory inference; pseudotemporal ordering; single-cell transcriptomics
highlights: |
  These are the highlights. 
---
\* Equal contribution

@@

library(dynalysis)
library(tidyverse)

figs <- list()
sfigs <- list()
```

# Introduction
<!-- Three paragraph introduction, typical for Nat Biotech papers -->
<!-- General into about single-cell & trajectories -->
Single-cell -omics technologies now make it possible to more accuratly model biological systems than ever before. One area where single-cell data can be particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation. When cells are sampled from a population in which cells are at different unknown points in the dynamic process, trajectory inference methods can be used to computationally order these cells along their dynamic process [@cannoodt_computational_2016]. Offering an unbiased and transcriptome-wide understanding of the dynamic process, trajectories then allow the identification new (primed) subsets of cells, delineation the exact wiring of a differentiation tree and inference of regulatory interaction responsible for a bifurcation <!-- refs -->.

<!-- Discussion of TI, history, different methods, brief future perspectives -->
Dozens of trajectory inference methods have been developed over the last years, and more are being published almost every month. At first, most trajectory inference was done for linear trajectories, but several methods which can detect complex branching or cyclical behaviour

<!-- Context and general overview of our evaluation -->
Given this plethora of available trajectory inference methods,

Trajectories can then be used to delineate the different stages in the dynamic process and their associated gene signatures <!-- REF -->,

# Methods

@@

<!-- I would move some stuff from this paragraph to the supplementary if necessary -->
We gathered a total of `r sum(method_df$is_ti == "Yup", na.rm=TRUE)` TI methods, and selected a subset of `r sum(!is.na(method_df$qc_score))` for evaluation, based on free availability and being able to be used within a programming context. We characterized the methods from both a computational and a user perspective. Computationally, methods use a variety of algorithms and subcomponents to reach a final ordering. However, most use a common framework consisting of three steps: (i) preprocessing (filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction or clustering and (iii) ordering the cells along the simplified representation. <!-- -> add supplementary information -->  We assessed the quality of the implementation of each method, using metrics looking at code availability, documentation, code style and extendability. We first contacted all the authors, to allow them to improve their method before publishing our evaluation. While most methods lack at several metrics, code quality tends to increase over time.

@@

>>>>

---
title: "An evaluation framework for trajectory inference methods"
author:
  - Robrecht Cannoodt*
  - Wouter Saelens*
  - Helena Todorov
  - Yvan Saeys
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
    fig_caption: yes
    reference_docx: templates/template.docx
bibliography: references.bib
csl: nature-biotechnology.csl
abstract: |
  Using single-cell transcriptomics data, it is now possible to computationally order cells along trajectories, allowing the unbiased transcriptome-wide study of cellular dynamic processes. Dozens of trajectory inference methods have been developed, early methods mostly focussing on linear trajectories, while recently methods also model more complex branching trajectories. However, the performance of these methods have not yet been comprehensively evaluated. Here, we compare a total of 22 trajectory inference methods, on both realistic synthetic datasets and a large set is real datasets. We employ a novel parameter optimisation procedure, and compare methods using several metrics, including accuracy of the inferred ordering, correctness of the network wiring, code quality and user friendliness. We conclude with practical guidelines for method users, and also believe our evaluation framework can be used to spearhead the development for new methods. Our evaluation pipeline is fully reproducible, can be easily extended and is available at www.github.com/dynverse.
keywords: |
  trajectory inference; pseudotemporal ordering; single-cell transcriptomics
highlights: |
  These are the highlights. 
---
\* Equal contribution

@@

library(dynalysis)
library(tidyverse)
library(cowplot)

# refering to figures & supplementary figures
refs <- tibble(ref_id = character(), name = character(), ref_type = character())
ref <- function(ref_type, ref_id, suffix="") {
  if(nrow(refs %>% filter(ref_id == !!ref_id)) == 0) {
    refs <<- refs %>% bind_rows(tibble(
      name = create_names[[ref_type]](sum(refs$ref_type == ref_type) + 1),
      ref_type = ref_type,
      ref_id = ref_id
    ))
  }
  refs %>% filter(ref_id == !!ref_id) %>% pull(name) %>% paste0("**", ., suffix, "**")
}
create_names <- list(
  sfig = function(i) {glue::glue("Supplementary Figure {i}")},
  fig = function(i) {glue::glue("Figure {i}")}
)
```

# Introduction
<!-- Three paragraph introduction, typical for Nat Biotech papers -->
<!-- General into about single-cell & trajectories -->
Single-cell -omics technologies now make it possible to more accuratly model biological systems than ever before. One area where single-cell data can be particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation. When cells are sampled from a population in which cells are at different unknown points in the dynamic process, trajectory inference methods can be used to computationally order these cells along their dynamic process [@cannoodt_computational_2016]. Offering an unbiased and transcriptome-wide understanding of the dynamic process, trajectories then allow the identification new (primed) subsets of cells, delineation the exact wiring of a differentiation tree and inference of regulatory interaction responsible for a bifurcation<!-- refs -->.

<!-- Discussion of TI, history, different methods, brief future perspectives -->
Dozens of trajectory inference methods have been developed over the last years, and more are being published almost every month (`r ref("sfig", "methods_over_time", "a")`). Initially most trajectory inference was focused on linear trajectories, with one start and one end point. However, the prospect of branching trajectories was already described early [@trapnell_dynamics_2014], and about half of the current methods can handle one or more bifurcations in the trajectory (`r ref("sfig", "methods_over_time", "b")`). Some methods can already handle more complex, cyclical and branching behaviour, and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes hapening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].

```{r, fig.width = 15, fig.height = 10}
n_methods_over_time <- readRDS(figure_file("n_methods_over_time.rds", experiment_name="method_characteristics"))
trajectory_components_over_time <- readRDS(figure_file("trajectory_components_over_time.rds", experiment_name="method_characteristics"))

cowplot::plot_grid(plotlist=list(n_methods_over_time, trajectory_components_over_time), nrow=2, rel_heights = c(0.7, 0.3), labels="auto")
```
`r ref("sfig", "methods_over_time")` New TI methods over time. **a** Number of methods published or in preprint. **b** Number of methods which can handle a particular type of trajectory.

<!-- Common framework of TI, different components, naturally progressing towards the need for an evaluation -->
Although trajectory inference methods use a variety of algorithms and subcomponents to reach a final ordering, most consist of the following three steps: (i) preprocessing (filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction or clustering and (iii) ordering the cells along the simplified representation. These components are frequently interchangeable. Depending on the way the cells are ordered, some prior information about the dynamic process can also be required, or optional. <!---- TODO --->

<!-- Context and general overview of our evaluation -->
Given this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal method available for their problem. Moreover, new methods need to be rigorously tested, so that development can focus on improving the current state-of-the-art. In this study, we therefore for the first time developed a comprehensive evaluation framework for trajectory inference methods. We test both on synthetic data, for which the gold standard is known, and real data, for which in some cases the gold standard can be extracted from current knowledge. We include a rigorous parameter optimisation. To make our framework reusable and extendable, we make use of continuous analysis [@beaulieu-jones_reproducibility_2017], which allows the developers of new methods to easily test their methods and compare them against the state-of-the-art.

@@

<!-- I would move some stuff from this paragraph to the supplementary if necessary -->
We gathered a total of `r sum(method_df$is_ti == "Yup", na.rm=TRUE)` TI methods, and selected a subset of `r sum(!is.na(method_df$qc_score))` for evaluation, based on free availability and being able to be used within a programming context. We characterized the methods from both a computational and a user perspective.  <!-- -> add supplementary information -->  We assessed the quality of the implementation of each method, using metrics looking at code availability, documentation, code style and extendability. We first contacted all the authors, to allow them to improve their method before publishing our evaluation. While most methods lack at several metrics, code quality tends to increase over time.

@@

# Online Methods
